---
title: "Statistical Inference Project"
output: pdf_document
---

##Part 1: Simulation Exercises

**Context: **
The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also also 1/lambda. Set lambda = 0.2 for all of the simulations. In this simulation, you will investigate the distribution of averages of 40 exponential(0.2)s. Note that you will need to do a thousand or so simulated averages of 40 exponentials.

We can start by generating the simulation data. To do this we will be taking the mean of 40 random observations from the exponential distribution with our rate parameter lambda set to 0.2. We will generate 1000 such observations and then use data from these observations to answer the questions in part 1. Below is the R code required to generate the data we will be using

```{r}

#seed is set to ensure reproducibility
set.seed(1234)

#Assign our rate parameter
lambda <- 0.2
n <- 40

#Create a dataframe to store out generated observation values
simData <- data.frame(simMean = numeric(0), simVariance = numeric(0), lowerInterval = numeric(0), upperInterval = numeric(0), inInterval = logical())

#Collect the means,variances,upper and lower intervals for 40 random observations in the exponential distribution 1000 times
for(i in 1:1000){
    
    rObservation <- rexp(n,lambda)
    
    trialMean <- mean(rObservation)
    CI <- trialMean + c(-1, 1) * 1.96 * sd(rObservation) / sqrt(n)
    
    simData[i,1]<- trialMean
    simData[i,2] <- var(rObservation)
    simData[i,3] <- CI[1]
    simData[i,4] <- CI[2]
    simData[i,5] <- (1/lambda) > CI[1] && (1/lambda) < CI[2]
    
}


#Display the first new observation means
head(simData, n=5)



```

**1. Show where the distribution is centered at and compare it to the theoretical center of the distribution.**

As was stated in the context section above the theoretical mean and the theoretical standard deviation of the exponential distribution should be 1/lambda. In our case this value is 5. So we will have to calculate the observed mean and standard deviation for our simulated data and see how they compare. 

In the plot below we have used a histogram to show the density of the observed means. We have also overlayed a normal distribution on the simulated data to show that is is approximately normal. As we can also see the theoretical mean is close to the observed mean.

```{r}

theoreticalMean <- 1/lambda
theoreticalMean

observedMean <- mean(simData$simMean)
observedMean

library(ggplot2)

dataPlot <- ggplot(simData,aes(x=simMean)) + geom_histogram(binwidth=.5,aes(y=..density..),colour="black",fill="white") + geom_vline(aes(xintercept=mean(simData$simMean)),color = "blue",linetype="dashed",size=1) + geom_text(aes(x=mean(simData$simMean)-0.2,label="Actual Mean"),angle=90, y=0.3)+ xlab("Mean of 40 exponentials") + ggtitle("Histogram of Simulated Data") + stat_function(fun=dnorm,colour = "black",arg = list(mean = 1/lambda,sd = 1/lambda*(1/sqrt(n)))) + geom_vline(aes(xintercept=1/lambda),color = "red",linetype="dashed",size=1) + geom_text(aes(x=(1/lambda)+0.2,label="Theoretical Mean"),angle=90, y=0.3)
                
dataPlot

```

**2. Show how variable it is and compare it to the theoretical variance of the distribution.**

To show how variable the distribution is we need to calculate its variance and then we should compare it to the theoretical variance. As we can see by the values calculated belows the values of the two quantities are close to each other.

```{r}

theoreticalVariance <- (1/lambda^2)/n
theoreticalVariance

observedVariance <- var(simData$simMean)
observedVariance


```

*3. Show that the distribution is approximately normal.*

It should be noted that a graphical demonstration of this approximate normality can be seen in the plot above on which we have overlayed the normal distribution. We will now discuss other factors that could help show the distribution is normal.

One way (and this is by no means the best way) is that since we know that the means and median will be equalin normally distributed data we can check the mean and median value of simulatedData we have generated. These values are shown below and appear to be approximately equal. 

```{r}
mean(simData$simMean)

median(simData$simMean)

```

We can also leverage the power of the CLT help us demonstrate that this distribution is approximately normal. The CLT states that the distribution of averages of iid variables (properly normalized) becomes that of a standard normal as the sample size increases. Considering that we have made 1000 observations we can expect our distribution to be fairly normal but as the data we collected limits to infinity we would expect closer approximations to the actual normal distribution. 

**4. Evaluate the coverage of the confidence interval for 1/lambda: X¯ ± 1.96 S/sqrt(n)**

Since earlier in our data analysis we had generated data such as the confidence intervals for each individual observation of 40 random exponentials and also evaluated if these individual confidence intervals either contained or did not contain the theoretical mean we are now in a very good position to evaluate the coverage of the confidence interval. 

We are able to calculate our percentage confidence by taking the TRUE count in the table below and dividing it by our number of observations which is in this case 1000. 
```{r}
#Shows us the counts of intervals that contain and dont contain the theoretical mean
result <- table(simData$inInterval)
result

print(paste(c("The coverage of the confidence interval is ", result[2]/1000),collapse = " "))

```